Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-01-30 paper/zhang-cook-laa-langrene-menendez.Rmd
section monotonicity

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE,
                      root.dir = here::here(), 
                      fig.path = here::here("figs/"), 
                      cache = TRUE,
                      fig.align = "center",
                      fig.height = 3,
                      out.width = "100%")
```

@@

@@

The goal of continuous optimisation is to find the best solution within the space of all feasible solutions where typically the best solution is decided by an objective function. Broadly speaking, optimization can be unconstrained or constrained [@kelley1999iterative]. The unconstrained problem can be formulated as a minimization (or maximization)  problem such as 
$\min_{x} f(x)$ where $f:\mathbb{R}^n \rightarrow \mathbb{R}$ is an objective function with  certain properties defined in an $L^p$ space. In this case, solutions rely on gradient descent or ascent methods. In the constrained optimization problem additional restrictions  are introduced via a set of functions that can be convex or non-convex: $g_i:\mathbb{R}^n \rightarrow \mathbb{R}$ for $i = 1, \ldots k$ and hence the problem can be written as
$\min_{x} f(x)$ *subject to* $g_i(x) \leq 0$. Here methods such as Langrange multipliers and convex optimization methods including linear and quadratic programming can be used.

@@

A starting point of diagnosing an optimiser is to understand how many search the optimiser has conducted. One may want to simply plot the index value of the search points across its natural order. A point geometry may work well if each iteration has a similar number of points,  however, the plot will be disproportionate if some iterations have considerably more points than other - it over-emphasizes the iterations that have more search points! An alternative is to summarise the search in each iteration using boxplot where each iteration will be spaced out equally. Occasionally, one may still want to switch back to point geometry if the number of points is small and this can be achieved via the `cutoff` argument. Additional annotation can be added with new layers, for example, the number of points searched in each iteration can be added as text label at the bottom of each iteration; the anchor bases to interpolate are connected and highlighted in a larger size; and the color of the last iteration is in a grey scale to indicate no better basis found in this iteration.

@@

```{r toy-search, fig.env = "widefigure", fig.cap="A comparison of search by two otimisers: SA (left) and SAJO (right) on a 2D projection of a six-variable dataset, \\code{boa6}, using the holes index. Both optimisers find the final basis with similar index value while it takes SAJO more iteration because the algorithm allows a probabilistic acceptance of bases with lower index value, as observed in iteration 4 and 6-8."}
```

@@

```{r toy-interp, fig.env = "widefigure", fig.cap = "Trace plots of the interpolating bases with the same optimisation routine as the previous figure. Both traces are smooth while the change of index value in each iteration may not be monotinic."}
```

@@

```{r toy-pca, fig.height=6, fig.cap = "The PCA plot of 1D projection problem on the 5-variable dataset, \\code{boa5}, using the holes indexwith two optimisers: SA and PD. All the bases in PD has been flipped and a grey dashed line has been annotated to indicate the symmetry of two starting bases."}
```

Apart from checking the progression of an optimiser, looking at how the points evaluated by the optimisers look like in the space is another interest. Given the orthonormality constraint, the space of projection bases $\mathbf{A}_{p \times d}$ is a $p \times d$ dimension sphere and  dimension reduction methods, i.e. principal component analysis is applied first to project all the bases onto a 2D space. In a projection pursuit guided tour optimisation, there are various type of bases involved: 1) The start basis; 2) The anchor bases that have the highest index value in each iteration; 3) The search bases that an optimiser evaluated to produce the anchor bases; 4) The interpolating bases on the interpolation path; and finally 5) the end basis. The importance of these basis differs with the most important ones being the start, interpolating and end bases. Anchor and search bases can be turned on with argument `details = TRUE`. Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign. This happens because the projection is invariate to the sign difference of the bases and so does the index value, however, this creates difficulties for comparing end bases. A flip sign device is implemented to flip the sign of all the bases in one optimisation if the situation described above happens and to ensure different routines finish close to each other.

Several annotations have been made to help understanding this plot. The original space of the bases is a high dimensional sphere and random bases on the sphere can be generated via the CRAN package \CRANpkg{geozoo} [@geozoo]. Along with the bases recorded during the optimisation and a zero basis, PCA is performed to get the first two PC coordinates of all the bases. The search space in the 2D space is a circle with the origin being the PC coordinates of the zero matrix, and radius estimated as the largest distance between the origin and all the bases. The theoretical best basis is known with simulated data and can be labelled for comparing the how close the ending basis to the theoretical one for different optimisers. Various aesthetics, i.e. size, alpha and color, are applicable to emphasize the crucial elements and adjust for the presentation. For example, anchor points and search points are less important and a smaller size and alpha is used; The interpolation paths are more informative if alpha increases from start to finish (that is, the path is more opaque towards the end).

@@

Animation is another display to show how the search porgresses from start to finish in the space and an  `animate = TRUE` argument is used to enable the animation. Figure \ref{fig:toy-pca-animated} shows six frames from the animation of the PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information one can learn from this animation is that the SA search finds its end basis quicker than the PD search since SA has finished in the 5th frame while PD is still making more progression.

@@

@@

For a particular index function, the best algorithm to optimise depends on the character of the index and the data. If the index function is smooth and has a single maximum, all of the three algorithms introduced above can find the maximum.  When multiple optima are present, simulated annealing (SA) may get stuck at a local maximum.  In the case where the index function is non-smooth, pseudo derivative search may even fail to find the maximum. In this section, examples will be presented to outline how the diagnostic plots can be used to compare the performance of optimisers.

@@

Random variables with different distributions have been simulated and the distributional form of each variable is presented in Equations \ref{eq:sim-norm} to \ref{eq:sim-x7}. Variable `x1`, `x8`, `x9` and `x10` are normal noise with zero mean and unit variance and `x2` to `x7` are normal mixtures with varied weights and locations. All the variables have been scaled to have an overall unit variance before running the projection pursuit.

@@

A non-monotonic interpolation path has already been given in Figure \ref{fig:toy-interp}: to get to the target basis from the current one, the interpolator may have passed a basis with even higher index value. This motivates the design of an interruption to check whether higher index value can be reached during the interpolation. If such a basis can be found, the optimiser will start the next iteration from that basis instead of the target one. Figure \ref{fig:interruption} contrasts the trace with and without the interruption for the simulated annealing optimiser and it can be observed that while the first two interpolation are identical, a better basis has been found than the target one in the third interpolation. Rather than starting the next iteration from the "target basis" on Time 62, as shown in the left panel, the interruption, on the right panel, starts the next iteration from the "interpolation basis" on Time 61. The interruption has also detected better basis on the fourth interpolation. With the interruption, the optimiser is able to find the final basis with a higher index value (0.865 vs. 0.879).

```{r interruption, fig.env = "widefigure", fig.cap = "Trace plot of 2D projection on \\code{boa6} data with holes index optimised by optimiser SA. The purpose of the plot is to compare the effect of implementing the interruption that avoid fully interpolating to the target basis if a better one can be found during the interpolation. With this implementation, the optimiser is able to find a better final basis."}
```

@@

## fliping the sign in the bases

@@

```{r kol-result, fig.height = 6, fig.cap="1D projection of data \\code{boa6} on \\code{norm\\_kol} index,  optimised by SA and SAJO. Two different neighbourhood size of 0.5 and 0.7 is used. The finish point of each path is larger than the start point and optimums are separately labelled with * representing the global optimum and x the local optimum The color of the path is based on whether the optimiser finds the global or local optimum, after polishing."}
```

@@

- \code{get\_*()} extracts and manipulates certain components from the existing data object.
  - \code{get\_best()} extracts the best basis found in the data object
  - \code{get\_start()} extracts the starting basis
  - \code{get\_interp()} extracts the observations in the interpolation
  - \code{get\_interp\_last()} extracts the end observations of the interpolation in each iteration
  - \code{get\_anchor()} extracts the target observations found by the optimiser
  - \code{get\_search()} extracts the search observations evaluated by the optimiser
  - \code{get\_search\_count()} produces the summary table of the number of observation in each iteration
  - \code{get\_center()} produces the center point of the basis space estimated by the starting points
  - \code{get\_space\_param()} produces the coordinates of the center and radius of the basis space
  - \code{get\_theo()} extracts the theoretical observations from the data object
  - \code{get\_interrupt()} extract the end point of the interpolation and the target point when an interruption happens
  - \code{get\_basis\_matrix()}: flattens all the bases into a matrix

@@

- Utilities
  - \code{add\_*()} are internal wrapper functions that facilitate the composition of PCA plot
  - \code{theme\_fern()} and \code{format\_label()} for better display of the grid lines and axis formatting
  - \code{clean\_method()} for clean up the name of the optimisers
  - \code{botanical\_palettes} is a collection of color palettes from Australian native plants. Quantitative palettes include daisy, banksia and cherry and sequential palettes contain fern and acacia.
  - \code{botanical\_pal()} as the color interpolator 
  - \code{scale\_color\_botanical()} is a ggplot2 scale for botanical palettes.

@@

This paper has illustrated setting up a data object that can be used for diagnosing a complex optimisation procedure. The ideas were illustrated using the optimisers available for projection pursuit guided tour. Here the constraint is the orthornormality condition of the projection bases. The approach used here could be broadly applied to understand other constrained optimisers.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE,
                      root.dir = here::here(), 
                      fig.path = here::here("figs/"), 
                      cache = TRUE,
                      fig.align = "centre",
                      fig.height = 3,
                      out.width = "100%")
```

@@

```{r }
# spelling::spell_check_files(here::here("paper/zhang-cook-laa-langrene-menendez.Rmd"), lang = "en-GB", ignore = spelling::get_wordlist())
```

@@

The goal of continuous optimisation is to find the best solution within the space of all feasible solutions where typically the best solution is decided by an objective function. Broadly speaking, optimization can be unconstrained or constrained [@kelley1999iterative]. The unconstrained problem can be formulated as a minimization (or maximization)  problem such as 
$\min_{x} f(x)$ where $f:\mathbb{R}^n \rightarrow \mathbb{R}$ is an objective function with  certain properties defined in an $L^p$ space. In this case, solutions rely on gradient descent or ascent methods. In the constrained optimization problem additional restrictions  are introduced via a set of functions that can be convex or non-convex: $g_i:\mathbb{R}^n \rightarrow \mathbb{R}$ for $i = 1, \ldots k$ and hence the problem can be written as
$\min_{x} f(x)$ *subject to* $g_i(x) \leq 0$. Here methods such as multipliers and convex optimization methods including linear and quadratic programming can be used.

@@

A starting point of diagnosing an optimiser is to understand how many search the optimiser has conducted. One may want to simply plot the index value of the search points across its natural order. A point geometry may work well if each iteration has a similar number of points,  however, the plot will be disproportionate if some iterations have considerably more points than other - it over-emphasizes the iterations that have more search points! An alternative is to summarise the search in each iteration using boxplot where each iteration will be spaced out equally. Occasionally, one may still want to switch back to point geometry if the number of points is small and this can be achieved via the `cutoff` argument. Additional annotation can be added with new layers, for example, the number of points searched in each iteration can be added as text label at the bottom of each iteration; the anchor bases to interpolate are connected and highlighted in a larger size; and the colour of the last iteration is in a grey scale to indicate no better basis found in this iteration.

@@

```{r toy-search, fig.cap="A comparison of search by two otimisers: SA (left) and SAJO (right) on a 2D projection of a six-variable dataset, \\code{boa6}, using the holes index. Both optimisers find the final basis with similar index value while it takes SAJO more iteration because the algorithm allows a probabilistic acceptance of bases with lower index value, as observed in iteration 4 and 6-8."}
```

@@

```{r toy-interp, fig.cap = "Trace plots of the interpolating bases with the same optimisation routine as the previous figure. Both traces are smooth while the change of index value in each iteration may not be monotinic."}
```

@@

```{r toy-pca, fig.height=5, fig.cap = "The PCA plot of 1D projection problem on the 5-variable dataset, \\code{boa5}, using the holes indexwith two optimisers: SA and PD. All the bases in PD has been flipped and a grey dashed line has been annotated to indicate the symmetry of two starting bases."}
```

Apart from checking the progression of an optimiser, looking at how the points evaluated by the optimisers look like in the space is another interest. Given the orthonormality constraint, the space of projection bases $\mathbf{A}_{p \times d}$ is a $p \times d$ dimension sphere and  dimension reduction methods, i.e. principal component analysis is applied first to project all the bases onto a 2D space. In a projection pursuit guided tour optimisation, there are various type of bases involved: 1) The start basis; 2) The anchor bases that have the highest index value in each iteration; 3) The search bases that an optimiser evaluated to produce the anchor bases; 4) The interpolating bases on the interpolation path; and finally 5) the end basis. The importance of these basis differs with the most important ones being the start, interpolating and end bases. Anchor and search bases can be turned on with argument `details = TRUE`. Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign. This happens because the projection is invariant to the sign difference of the bases and so does the index value, however, this creates difficulties for comparing end bases. A flip sign device is implemented to flip the sign of all the bases in one optimisation if the situation described above happens and to ensure different routines finish close to each other.

Several annotations have been made to help understanding this plot. The original space of the bases is a high dimensional sphere and random bases on the sphere can be generated via the CRAN package \CRANpkg{geozoo} [@geozoo]. Along with the bases recorded during the optimisation and a zero basis, PCA is performed to get the first two PC coordinates of all the bases. The search space in the 2D space is a circle with the origin being the PC coordinates of the zero matrix, and radius estimated as the largest distance between the origin and all the bases. The theoretical best basis is known with simulated data and can be labelled for comparing the how close the ending basis to the theoretical one for different optimisers. Various aesthetics, i.e. size, alpha and colour, are applicable to emphasize the crucial elements and adjust for the presentation. For example, anchor points and search points are less important and a smaller size and alpha is used; The interpolation paths are more informative if alpha increases from start to finish (that is, the path is more opaque towards the end).

@@

Animation is another display to show how the search progresses from start to finish in the space and an  `animate = TRUE` argument is used to enable the animation. Figure \ref{fig:toy-pca-animated} shows six frames from the animation of the PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information one can learn from this animation is that the SA search finds its end basis quicker than the PD search since SA has finished in the 5th frame while PD is still making more progression.

@@

Something to fill here ...

@@

In this section, several examples will be presented to show how the diagnostic plots discover something unexpected in the optimisation algorithm, and guide the implementation of some new features.

@@

Random variables with different distributions have been simulated and the distributional form of each variable is presented in Equations \ref{eq:sim-norm} to \ref{eq:sim-x7}. Variable `x1`, `x8`, `x9` and `x10` are normal noise with zero mean and unit variance, and `x2` to `x7` are normal mixtures with varied weights and locations. All the variables have been scaled to have an overall unit variance before projection pursuit.

@@

An example of non-monotonic interpolation has been given in Figure \ref{fig:toy-interp}: a path that passes bases with higher index value than the target one. With optimiser SAJO, a non-monotonic interpolation is justified since the target bases do not necessarily have a higher index value than the current one, while this is not the case for SA. The original trace plot for a 2D projection, optimised by SA, is shown on the left panel of Figure \ref{fig:interruption} and one can observe clearly that the non-monotonic interpolation has undermined the optimiser to realise its full potential. Hence, an interruption is implemented to stop the interpolation at the basis with the highest index value. The right panel of Figure \ref{fig:interruption} presents the trace plot after implementing such interruption and it can be observed that while the first two interpolations are identical, in the third interpolation the basis at time 61 has a higher index value than the target. Rather than starting the next iteration from the target basis on Time 65 as shown in the left panel, the optimiser, on the right panel, starts the next iteration from the interpolation basis on Time 61. With the interruption, the optimiser is able to find the final basis with a index value higher than the most promising basis when the interruption is not implemented.

```{r interruption, fig.cap = "Trace plot of 2D projection on \\code{boa6} data with holes index, optimised by optimiser SA."}
```

@@

## Flipping the sign in the bases

@@

```{r kol-result, fig.height = 5, fig.cap="1D projection of data \\code{boa6} on \\code{norm\\_kol} index,  optimised by SA and SAJO. Two different neighbourhood size of 0.5 and 0.7 is used. The finish point of each path is larger than the start point and optimums are separately labelled with * representing the global optimum and x the local optimum The colour of the path is based on whether the optimiser finds the global or local optimum, after polishing."}
```

@@

- \code{get\_*()} extracts and manipulates certain components from the existing data object.
  - \code{get\_best()} extracts the best basis found in the data object
  - \code{get\_start()} extracts the starting basis
  - \code{get\_interp()} extracts the observations in the interpolation
  - \code{get\_interp\_last()} extracts the end observations of the interpolation in each iteration
  - \code{get\_anchor()} extracts the target observations found by the optimiser
  - \code{get\_search()} extracts the search observations evaluated by the optimiser
  - \code{get\_search\_count()} produces the summary table of the number of observation in each iteration
  - \code{get\_centre()} produces the centre point of the basis space estimated by the starting points
  - \code{get\_space\_param()} produces the coordinates of the centre and radius of the basis space
  - \code{get\_theo()} extracts the theoretical observations from the data object
  - \code{get\_interrupt()} extract the end point of the interpolation and the target point when an interruption happens
  - \code{get\_basis\_matrix()}: flattens all the bases into a matrix

@@

- Utilities
  - \code{add\_*()} are internal wrapper functions that facilitate the composition of PCA plot
  - \code{theme\_fern()} and \code{format\_label()} for better display of the grid lines and axis formatting
  - \code{clean\_method()} for clean up the name of the optimisers
  - \code{botanical\_palettes} is a collection of colour palettes from Australian native plants. Quantitative palettes include daisy, banksia and cherry and sequential palettes contain fern and acacia.
  - \code{botanical\_pal()} as the colour interpolator

@@

This paper has illustrated setting up a data object that can be used for diagnosing a complex optimisation procedure. The ideas were illustrated using the optimisers available for projection pursuit guided tour. Here the constraint is the orthonormality condition of the projection bases. The approach used here could be broadly applied to understand other constrained optimisers.
